---
author: "Mira Terdiman"
title: "252 Final Project Analysis"
---

```{r}
rm(list = ls()) #clear environment to make sure there are no overwriting errors
set.seed(252)
```

The purpose of this document is to compile all of our estimator methods and data (to avoid accidental conflicts/overwriting) and also include statistical inference for these methods. Each section will include the ATE for a particular estimator, as well as statistical inference for that ATE. The inference will be conducted via non-parametric bootstrapping.

To make the bootstrapping easier, I am turning each estimation process into a function that can then be applied repeatedly to our data (all functions assume data has already been cleaned to include variables A, Y, W1, etc). 

```{r, echo=FALSE, include=FALSE}
#loading necessary libraries
library(tidyr)
library(dplyr)
library(readr)
library(SuperLearner)
library(arm)
library(earth)
library(ltmle)
```


```{r, echo=FALSE}
clean_state_data <- read.csv("data/clean_state_data.csv")

#adding in/imputing missing data
clean_state_data$phf_per_capita[is.na(clean_state_data$phf_per_capita) & clean_state_data$state == "Georgia"] <- 108
clean_state_data$phf_per_capita[is.na(clean_state_data$phf_per_capita) & clean_state_data$state == "Mississippi"] <- 567831325/2954160
clean_state_data$phf_per_capita[is.na(clean_state_data$phf_per_capita)] <- median(clean_state_data$phf_per_capita, na.rm = TRUE)

#creating model data
model_df <- clean_state_data %>%
  dplyr::select(cases_per_100k,exposure_binary,pct_urban,pct_insuredE,pct_bachelors_plusE,pct_povertyE,phf_per_capita)|>
  rename(Y = cases_per_100k,
         A = exposure_binary,
         W1 = pct_urban,
         W2 = pct_insuredE,
         W3 = pct_bachelors_plusE,
         W4 = pct_povertyE,
         W5 = phf_per_capita)
```


# Unadjusted Estimator

## Estimator

Our unadjusted estimator is simply taking the expected difference in our outcome $Y$ across our two exposure levels, $A = 0,1$. 

```{r}
ATE_un <- function(dat){
  ATE_unadj <- mean(dat$Y[dat$A == 1]) -
             mean(dat$Y[dat$A == 0])
  return(ATE_unadj)
}

ATE_un(model_df)
```

Our unadjusted ATE is 0.420. Specifically, the states with more permissive vaccine exemption policies had, on average, about 0.420 additional measles cases per 100,000 people or 4.2 additional measles cases per 10,000 people compared to states with stricter policies. This is a naive estimator, and we will now run the three techniques we learned in class.

## Inference

```{r}
B <- 500

unadj_boot_est <- data.frame(matrix(NA, nrow = B, ncol = 1))

for(b in 1:B){
  n <- nrow(model_df)
  index.boot <- sample(1:n, replace = T)
  dt_boot <- model_df[index.boot,]
  
  unadj_boot_est[b,] <- ATE_un(dat = dt_boot)
}
colnames(unadj_boot_est) <- "Unadjusted Estimator"

summary(unadj_boot_est$`Unadjusted Estimator`)
var(unadj_boot_est$`Unadjusted Estimator`)
```

```{r}
library(ggplot2)
unadj_boot_est %>%
  ggplot(aes(x = `Unadjusted Estimator`)) +
  geom_histogram() +
  theme_minimal() +
  geom_vline(xintercept = mean(unadj_boot_est$`Unadjusted Estimator`), color = "maroon")
```


# G-Computation

We will now perform G-Computation, both with and without using a super learner.

## Simple G-Computation

```{r}
g_comp_simp <- function(dat){
  gcomp_mod <- lm(Y ~ A + W1 + W2 + W3 + W4 + W5, data = dat)
  
  #defining potential outcomes
  data_A1 <- dat
  data_A1$A <- 1
  
  data_A0 <- dat
  data_A0$A <- 0
  
  #predictions
  Y1_pred <- predict(gcomp_mod, newdata = data_A1)
  Y0_pred <- predict(gcomp_mod, newdata = data_A0)
  
  ATE <- mean(Y1_pred - Y0_pred)
  return(ATE)
  
}

g_comp_simp(model_df)
```

Our adjusted ATE via G-Comp is 0.416. Specifically, the states with more permissive vaccine exemption policies had, on average, about 0.416 additional measles cases per 100,000 people or 4.16 additional measles cases per 10,000 people compared to states with stricter policies. This estimate is very similar to the unadjusted ATE. 

## G-Computation With Super Learner

```{r}
g_comp_sl <- function(dat, SL.library){
  
  #creating datasets
  X <- subset(dat, select = -Y)
  X1 <- X0 <-X
  X1$A <- 1 
  X0$A <- 0
  
  #modeling
  gcomp_mod_sl <- SuperLearner(Y = dat$Y,
                         X = X,
                         SL.library = SL.library,
                         cvControl = list(V = nrow(dat)),
                         family = "gaussian")
  
  #potential outcomes
  pred_y1 <- predict(gcomp_mod_sl, newdata = X1)$pred
  pred_y0 <- predict(gcomp_mod_sl, newdata = X0)$pred
  
  #ATE
  ATE_SL <- mean(pred_y1 - pred_y0)
  return(ATE_SL)
  
}


SL.library <- c( "SL.mean", "SL.glm", "SL.bayesglm", "SL.earth", "SL.gam")

g_comp_sl(model_df, SL.library)

```

## Statistical Inference

(doing simple G-comp, since using a superlearner gave a really weird result)

```{r}
B <- 500

gcomp_boot_est <- data.frame(matrix(NA, nrow = B, ncol = 1))

for(b in 1:B){
  n <- nrow(model_df)
  index.boot <- sample(1:n, replace = T)
  dt_boot <- model_df[index.boot,]
  
  gcomp_boot_est[b,] <- g_comp_simp(dat = dt_boot)
  
}
colnames(gcomp_boot_est) <- "GComp Estimator"

summary(gcomp_boot_est$`GComp Estimator`)
var(gcomp_boot_est$`GComp Estimator`)
```


# Inverse Probability Weighting

## Simple IPW

```{r}
IPW_simp <- function(dat){
  propensity_score <- glm(A ~ W1 + W2 + W3 + W4 + W5, family = 'binomial', data = dat)
  
  prob_exposed <- predict(propensity_score, type = 'response')
  prob_unexposed <- 1 - prob_exposed
  
  wt_exposed <- as.numeric(model_df$A == 1) / prob_exposed
  wt_unexposed <- as.numeric(model_df$A == 0) / prob_unexposed

# evaluate the IPW estimand
  ipw <- mean(wt_exposed*model_df$Y) - mean(wt_unexposed*model_df$Y)
  ipw

# evaluate the stabilized IPW estimand
  ipw_stabilized <- mean(wt_exposed*model_df$Y)/mean(wt_exposed) - mean(wt_unexposed*model_df$Y)/mean(wt_unexposed)
  ipw_stabilized   
  
  ipws <- list(unstabilixed = ipw, stabilized = ipw_stabilized)
  
  return(ipws)
  
}

IPW_simp(model_df)
```

## IPW With Super Learner

```
IPW_sl <- function(dat, SL.library){
  SL_propensity <- SuperLearner(
  Y = dat$A,
  X = subset(dat, select = -c(A,Y)),
  SL.library = SL.library,
  #cvControl = list(V=10, stratifyCV = TRUE),
  family = binomial())
  
  #predicted probabilities
  prob_exposed <- SL_propensity$SL.predict
  prob_unexposed <- 1 - prob_exposed
  
  # look for positivity violations
  # print(summary(prob_exposed))    
  # print(summary(prob_unexposed))
  
  # create weights
  wt_exposed <- as.numeric(model_df$A == 1) / prob_exposed
  wt_unexposed <- as.numeric(model_df$A == 0) / prob_unexposed
  
  # check weights
  # print(summary(wt_exposed))
  # print(summary(wt_unexposed))
  
  ipw <- mean(wt_exposed*model_df$Y) - mean(wt_unexposed*model_df$Y)
  
  ipw_stabilized <- mean(wt_exposed*model_df$Y)/mean(wt_exposed) - mean(wt_unexposed*model_df$Y)/mean(wt_unexposed)
  
  ipws <- list(unstabilixed = ipw, stabilized = ipw_stabilized)
  
  return(ipws)
  
}

IPW_sl(model_df, SL.library)
```

## Statistical Inference

### For Simple IPW

```{r}
B <- 500

ipw_simp_boot_est <- data.frame(matrix(NA, nrow = B, ncol = 1))

for(b in 1:B){
  n <- nrow(model_df)
  index.boot <- sample(1:n, replace = T)
  dt_boot <- model_df[index.boot,]
  
  ipw_simp_boot_est[b,] <- IPW_simp(dat = dt_boot)$stabilized
  
}
colnames(ipw_simp_boot_est) <- "IPW Estimator"

summary(ipw_simp_boot_est$`IPW Estimator`)
var(ipw_simp_boot_est$`IPW Estimator`)
```


### For Super Learner IPW

```
B <- 500
#had to majorly simplify the library:)
SL.library <- c( "SL.mean", "SL.glmnet")

ipw_boot_est <- data.frame(matrix(NA, nrow = B, ncol = 1))

for(b in 1:B){
  n <- nrow(model_df)
  index.boot <- sample(1:n, replace = T)
  dt_boot <- model_df[index.boot,]
  
   if (length(unique(dt_boot$A)) < 2) {
    ipw_boot_est[b,] <- NA
    next
  }
  
  #getting lots of warnings and model is not convering
  ipw_boot_est[b,] <- tryCatch(
  suppressWarnings(
    IPW_sl(dat = dt_boot, SL.library = SL.library)$stabilized
  ),
  error = function(e) NA
)
  
}
colnames(ipw_boot_est) <- "IPW Estimator"

summary(ipw_boot_est$`IPW Estimator`)
var(ipw_boot_est$`IPW Estimator`)
```


# TMLE

## Estimator

```{r}
# creating wrapper function for LOOCV

run_ltmle <- function(dat, W=NULL, A, Y, SL.library=NULL, verbose=F){
  df <- dat[,c(W,A,Y)]
  attr(SL.library, "return.fit") <- TRUE
  out <-  suppressWarnings(suppressMessages( 
    ltmle(data = df, Anodes = A, Ynodes = Y, 
          abar = list(1,0), 
          SL.library=SL.library,
          SL.cvControl =list(V=nrow(dat)), #leave-one-out CV
          variance.method='ic',
          estimate.time=F)))
  print('Summary of pscore')
  print(summary(out$cum.g))
  
  print(summary(out))
  # print(out$fit) # provides the TMLE fits (including SL)
  
  out 
}
```

```{r}
tmle_1 <- run_ltmle(dat = model_df, 
                    W = c("W1","W2","W3","W4","W5"), 
                    A = "A", 
                    Y = "Y", 
                    SL.library=SL.library
                    )

summary(tmle_1)$effect.measures$ATE$estimate
# get estimates & CIs
# treatment_est <- summary(tmle_1)$effect.measures$treatment$estimate
# control_est <- summary(tmle_1)$effect.measures$control$estimate
# control_ci_lower <- summary(tmle_1)$effect.measures$control$CI[1]
# control_ci_upper <- summary(tmle_1)$effect.measures$control$CI[2]
# treatment_ci_lower <- summary(tmle_1)$effect.measures$treatment$CI[1]
# treatment_ci_upper <- summary(tmle_1)$effect.measures$treatment$CI[2]
```

## Statistical Inference

```{r}
B <- 500
SL.library <- c( "SL.mean", "SL.glm", "SL.bayesglm", "SL.earth", "SL.gam")

TMLE_boot_est <- data.frame(matrix(NA, nrow = B, ncol = 1))

for(b in 1:B){
  n <- nrow(model_df)
  index.boot <- sample(1:n, replace = T)
  dt_boot <- model_df[index.boot,]
  
  fit <- suppressWarnings(
      run_ltmle(
        dat = dt_boot,
        W = c("W1","W2","W3","W4","W5"),
        A = "A",
        Y = "Y",
        SL.library = SL.library
      )
    )
  
  TMLE_boot_est[b,] <- summary(fit)$effect.measures$ATE$estimate
  

}
colnames(TMLE_boot_est) <- "TMLE Estimator"

summary(TMLE_boot_est$`TMLE Estimator`)
var(TMLE_boot_est$`TMLE Estimator`)
```

# Summary

```{r}
#creating large df with all results
inference_df <- bind_cols(gcomp_boot_est, ipw_simp_boot_est, TMLE_boot_est, unadj_boot_est)

inference_df_long <- inference_df %>%
  pivot_longer(
    cols = c(`GComp Estimator`, `IPW Estimator`, `TMLE Estimator`, `Unadjusted Estimator`),
    names_to = "Estimate",
    values_to = "ATE"
  )

inference_df_long %>%
  ggplot(aes(x = ATE)) +
  geom_histogram(color = "#06768d", fill = "turquoise") +
  facet_grid(~Estimate) +
  labs(y = "Frequency", x = "ATE Estimate") +
  theme_minimal() +
  ggtitle("Histogram of point estimates from four estimators over 500 bootstrapped samples")
```

```{r}
inference_df%>%
  ggplot(aes(x = `Unadjusted Estimator`)) +
  geom_histogram(color = "#06768d", fill = "turquoise") +
  labs(y = "Frequency", x = "Unadjusted ATE Estimate") +
  theme_minimal() +
  ggtitle("Point estimates from unadjusted over 500 bootstrapped samples")

inference_df%>%
  ggplot(aes(x = `GComp Estimator`)) +
  geom_histogram(color = "#06768d", fill = "turquoise") +
  labs(y = "Frequency", x = "G-Computation ATE Estimate") +
  theme_minimal() +
  ggtitle("Point estimates from G-Comp estimator over 500 bootstrapped samples")

inference_df%>%
  ggplot(aes(x = `IPW Estimator`)) +
  geom_histogram(color = "#06768d", fill = "turquoise") +
  labs(y = "Frequency", x = "IPW ATE Estimate") +
  theme_minimal() +
  ggtitle("Point estimates from IPW estimator over 500 bootstrapped samples")

inference_df%>%
  ggplot(aes(x = `TMLE Estimator`)) +
  geom_histogram(color = "#06768d", fill = "turquoise") +
  labs(y = "Frequency", x = "TMLE ATE Estimate") +
  theme_minimal() +
  ggtitle("Point estimates from TMLE estimator over 500 bootstrapped samples")
```

## 95% CIs

```{r}
create.CI <- function(pt, boot, alpha=0.05){
  Zquant <- qnorm(alpha/2, lower.tail=F)
  CI.normal <- c(pt - Zquant*sd(boot), pt + Zquant*sd(boot))
  CI.quant <- quantile(boot, prob=c(0.025,0.975))
  
  yay<- data.frame(t(c(pt, CI.normal, CI.quant)))
  colnames(yay)<- c('pt', paste0(c('CI.lo', 'CI.hi'), '.normal'),
                    paste0(c('CI.lo', 'CI.hi'), '.quant') )
  return(yay)
}

#simple substitution
ss<- create.CI(pt = 0.4204447, boot=inference_df[,"Unadjusted Estimator"])

#g-comp estimator
gc <- create.CI(pt = 0.4160642, boot=inference_df[,"GComp Estimator"])

#IPW estimator
ip <- create.CI(pt = 0.4247156, boot=inference_df[,"IPW Estimator"])

#TMLE estimator
tm <- create.CI(pt = summary(tmle_1)$effect.measures$ATE$estimate, boot=inference_df[, "TMLE Estimator"])

CI_95 <- bind_rows(ss, gc, ip, tm)


CI_95 <- CI_95 %>%
  mutate(Estimator = c("Unadjusted", "G-Computation", "IPW", "TMLE"))

as_tibble(CI_95)
```

