---
author: "Mira Terdiman"
title: "252 Final Project Analysis"
---

Using this doc to run the negative binomial model we discussed with case counts.

```{r}
#getting data set up
rm(list = ls())
set.seed(252)
library(tidyr)
library(ggplot2)

clean_state_data <- read.csv("data/clean_state_data.csv")

#adding in/imputing missing data
clean_state_data$phf_per_capita[is.na(clean_state_data$phf_per_capita) & clean_state_data$state == "Georgia"] <- 108
clean_state_data$phf_per_capita[is.na(clean_state_data$phf_per_capita) & clean_state_data$state == "Mississippi"] <- 567831325/2954160
clean_state_data$phf_per_capita[is.na(clean_state_data$phf_per_capita)] <- median(clean_state_data$phf_per_capita, na.rm = TRUE)
```


## G-comp with the other model:

```{r}
# 1. Fit Poisson outcome model with population offset
counts_fit <- glm(
  measles_cases ~ exposure_binary + pct_urban + pct_insuredE +
    pct_bachelors_plusE + pct_povertyE + phf_per_capita +
    offset(log(population)),
  family = poisson(),
  data = clean_state_data
)

# 2. Create counterfactual datasets
data_1 <- clean_state_data
data_1$exposure_binary <- 1

data_0 <- clean_state_data
data_0$exposure_binary <- 0

# 3. Predict counterfactual outcomes
mu1 <- predict(counts_fit, newdata = data_1, type = "response")
mu0 <- predict(counts_fit, newdata = data_0, type = "response")

# 4. Simple substitution (g-computation) ATE
ATE <- mean(mu1 - mu0)

# 5. Output
ATE
```

The average difference in expected number of measles cases per state if every state were exposed vs if no state were exposed, holding population structure and covariates fixed. On average across states, exposure is associated with an estimated change of 28.5 measles cases per state, after adjusting for population size and state-level covariates.

## Inference (Bootstrapping)

```{r}
counts_gcomp <- function(dat){
  
  mod <- glm(measles_cases ~ exposure_binary + pct_urban + pct_insuredE +
    pct_bachelors_plusE + pct_povertyE + phf_per_capita +
    offset(log(population)),
    family = poisson(),
    data = dat)
  
  #defining potential outcomes
  data_1 <- dat
  data_1$exposure_binary <- 1
  
  data_0 <- dat
  data_0$exposure_binary <- 0
  
  #predictions
  mu1 <- predict(mod, newdata = data_1, type = "response")
  mu0 <- predict(mod, newdata = data_0, type = "response")
  
  ATE <- mean(mu1 - mu0)
  return(ATE)
  
}

counts_gcomp(clean_state_data)


#running bootstrap

B <- 500

counts_boot_est <- data.frame(matrix(NA, nrow = B, ncol = 1))

for(b in 1:B){
  n <- nrow(clean_state_data)
  index.boot <- sample(1:n, replace = T)
  dt_boot <- clean_state_data[index.boot,]
  
  counts_boot_est[b,] <- counts_gcomp(dat = dt_boot)
  
}
colnames(counts_boot_est) <- "Count Estimator"

summary(counts_boot_est$`Count Estimator`)
var(counts_boot_est$`Count Estimator`)
```

### Histogram and 95% CI

```{r}
create.CI <- function(pt, boot, alpha=0.05){
  Zquant <- qnorm(alpha/2, lower.tail=F)
  CI.normal <- c(pt - Zquant*sd(boot), pt + Zquant*sd(boot))
  CI.quant <- quantile(boot, prob=c(0.025,0.975))
  
  yay<- data.frame(t(c(pt, CI.normal, CI.quant)))
  colnames(yay)<- c('pt', paste0(c('CI.lo', 'CI.hi'), '.normal'),
                    paste0(c('CI.lo', 'CI.hi'), '.quant') )
  return(yay)
}

CI <- create.CI(pt = 28.49236, boot=counts_boot_est[,"Count Estimator"])

CI

counts_boot_est%>%
  ggplot(aes(x = `Count Estimator`)) +
  geom_histogram(color = "#06768d", fill = "turquoise") +
  labs(y = "Frequency", x = "G-Comp ATE Estimate (Counts)") +
  theme_minimal() +
  ggtitle("Point estimates from G-Comp estimator over 500 bootstrapped samples")
```




